A Discrepancy-based Framework to Compare Robustness between Multi-Attribute Evaluations

### 2016-04-16 Docs/paper/paper.tex
paper_v1

% first the title is needed
\title{A Discrepancy-based Framework to Compare Robustness between Multi-Objective Evaluations}

@@

\begin{abstract}
Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives. In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts. However, an evaluation makes no sense without its robustness being given (in the ill-defined sense of reliability). Statistical robustness computation methods are highly dependent of underlying statistical models and a model-independent generic framework is still lacking. We propose a generic formulation of such a framework in the case of integrated aggregated indicators, that allows to define a relative measure of robustness taking into account data structure and indicator values. We implement and apply it to virtual data-driven cases of urban systems for Paris districts. First numerical results show the potentialities of this new method. Furthermore, its relative independence of system type and system model appears as an advantage compared to classical statistical robustness methods.
\keywords{Multi-objective Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
\end{abstract}

@@

Multi-objective problems are organically linked to the complexity of underlying systems. Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration always leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of physical, biological or social system that present emergence and self-organization properties, where objectives can e.g. be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system. The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design or multi-objective evaluation of an existing system ; our examples being in the second case but the overall method staying rather generic).}. The new notion of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where sustainability implies contradictory objectives. The example of transportation systems, which conception shifted during the second half of the 20th century from cost-benefit analysis to multi-criteria decision-making, is also typical of such systems~\cite{bavoux2005geographie}. Geographical system are now well studied from such a point of view in particular thanks to the integration of multi-objective frameworks within Geographical Information Systems~\cite{carver1991integrating}. As for the micro-case of eco-districts, meso and macro urban planning and design may be made sustainable through indicators evaluation~\cite{jegou2012evaluation}.

A crucial aspect of an evaluation is a certain notion of its reliability, that we call in a fuzzy way \emph{robustness}. Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework. Statistics naturally include such a notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}. The first example that comes to mind is the application of the law of large numbers to obtain the \emph{p-value} of a model fit, that can be interpreted as a confidence measure of estimates. Besides, confidence intervals and \emph{beta-power} are other important indicators of statistical robustness. Concerning multi-objective optimization, in particular through heuristic algorithms (for example genetic algorithms, or operational research solvers), the notion of robustness of a solution concerns more the stability of the solution on the phase space of the corresponding dynamical system. Recent progresses have been done towards unified formulation of robustness for a multi-objective optimization problem, such as~\cite{deb2006introducing} where robust Pareto-front as defined as solutions that are insensitive to small perturbations. In~\cite{1688537}, the notion of degree of robustness is introduced, formalized as a sort of continuity of other solutions in successive neighborhood of a solution.

However, there still lack a generic method to estimate robustness of an evaluation that would be model-independent, i.e. that would be extracted from data structure and indicators but that would not depend on the method used. Some advantages could be for example an \emph{a priori} estimation of potential robustness of an evaluation and thus to decide if the evaluation is worth doing. We propose here a framework answering this issue. It is data-driven and not model-driven in the sense that robustness estimation does not depend on how indicators are computed, as soon as they respect some assumptions that will be detailed in the following.

@@

Our approach needs to restrict the scope of application. We assume that objectives can be expressed as spatial integrals, so it should apply to any territorial system and our application cases are urban systems. It is not much restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size. The example we use in the following such as conditional means or sums suit well the assumption. Even an already spatially aggregated indicator can be interpreted as a spatial indicator by using a Dirac distribution on the centroid of the corresponding area.

@@

A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives. If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution. It is analog to aggregated utility techniques in economics and many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}. The subtlety lies in the choice of weights, i.e. the shape of the projection function, and various approaches have been developed to find weights depending on the nature of the problem. Recent work~\cite{dobbie2013robustness} proposed to compare robustness of different aggregation techniques through sensitivity analysis, performed by Monte-Carlo simulations on synthetic data. Distribution of biases where obtained for various techniques and some showed to perform significantly better than others. Robustness assessment still depended on models used in that work.

@@

The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results including sensitivity analysis ; section 4 finally discuss limitations and potentialities of the method.

@@

We describe now the abstract framework allowing theoretically to compare robustnesses of evaluations of two different urban systems. Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context. Intuitively, it relies on the same simple empirical base which results from the following statements :
\begin{itemize}
\item Urban systems can be seen from the information available, i.e. raw data describing the system. As a data-driven approach, this raw data is the basis of our framework and robustness will be determined by its structure.
\item From data are computed indicators (objective functions). We assume that a choice of indicators is a will to translate particular aspects of the system, that is to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
\item Given many systems and associated indicators, a common space can be built to compare themselves. In that space, data will represent more or less well real systems, depending e.g. on initial scale, precision of data, missing data. We precisely capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory which express how a dataset is well distributed in the space it is embedded in~\cite{dick2010digital}. Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
\end{itemize}

%%%%%%%%%%%%%%%%
\subsection{Indicators}

@@

%%%%%%%%%%%%%%%%
\subsection{Aggregation}

As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation. Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method. Let define weights for the linear aggregation. We assume the indicators normalized, i.e. $q_c \in [0,1]$, for a more simple construction of relative weights.

Weights result of two components, one deterministic and one ``subjective'' part permitting more flexibility and the inclusion of judgement in the decision process. For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is a mean of :

\begin{itemize}
\item objective weight, corresponding to relative importance of the indicator
\[w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}\]
where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e. the effectively calculated value)
\item subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight. In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.
\end{itemize}

%%%%%%%%%%%%%%%%
\subsection{Robustness Estimation}

@@

@@

We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework. It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.

%%%%%%%%%%%%%%%%
\subsection{Data Collection}

We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris. We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features. We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}. Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source. An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.

%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/ex_data}
\caption{Example of a visualization of used data for center area of Paris. We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green. Dataset overall consists of around 200000 building features and 100000 road segments.}
\label{fig:data_ex}
\end{figure}
%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
\subsection{Implementation}

@@

%%%%%%%%%%%%%%%%
\subsection{Virtual Cases}

@@

\begin{table}

\begin{tabular}[10pt]{c|c|c|c|c}
Arrdt & $<\bar{d}_w> \pm \sigma (\bar{d}_w)$ & $<\bar{\varphi}> \pm \sigma (\bar{\varphi})$ & $<\bar{p}> \pm \sigma (\bar{p})$ & $R_{i,1}$ \\[3pt]
\hline
1 th & 0.731655 $\pm$ 0.041099 & 0.917462 $\pm$ 0.026637 & 0.191615 $\pm$ 0.052142 & 1.000000 $\pm$ 0.000000\\[3pt]
\hline
2 th & 0.723225 $\pm$ 0.032539 & 0.844350 $\pm$ 0.036085 & 0.209467 $\pm$ 0.058675 & 1.002098 $\pm$ 0.039972\\[3pt]
\hline
3 th & 0.713716 $\pm$ 0.044789 & 0.797313 $\pm$ 0.057480 & 0.185541 $\pm$ 0.065089 & 0.999341 $\pm$ 0.048825\\[3pt]
\hline
4 th & 0.712394 $\pm$ 0.042897 & 0.861635 $\pm$ 0.030859 & 0.201236 $\pm$ 0.044395 & 0.973045 $\pm$ 0.036993\\[3pt]
\hline
5 th & 0.715557 $\pm$ 0.026328 & 0.894675 $\pm$ 0.020730 & 0.209965 $\pm$ 0.050093 & 0.963466 $\pm$ 0.040722\\[3pt]
\hline
6 th & 0.733249 $\pm$ 0.026890 & 0.875613 $\pm$ 0.029169 & 0.206690 $\pm$ 0.054850 & 0.990676 $\pm$ 0.031666\\[3pt]
\hline
7 th & 0.719775 $\pm$ 0.029072 & 0.891861 $\pm$ 0.026695 & 0.209265 $\pm$ 0.041337 & 0.966103 $\pm$ 0.037132\\[3pt]
\hline
8 th & 0.713602 $\pm$ 0.034423 & 0.931776 $\pm$ 0.015356 & 0.208923 $\pm$ 0.036814 & 0.973975 $\pm$ 0.033809\\[3pt]
\hline
9 th & 0.712441 $\pm$ 0.027587 & 0.910817 $\pm$ 0.015915 & 0.202283 $\pm$ 0.049044 & 0.971889 $\pm$ 0.035381\\[3pt]
\hline
10 th & 0.713072 $\pm$ 0.028918 & 0.881710 $\pm$ 0.021668 & 0.210118 $\pm$ 0.040435 & 0.991036 $\pm$ 0.038942\\[3pt]
\hline
11 th & 0.682905 $\pm$ 0.034225 & 0.875217 $\pm$ 0.019678 & 0.203195 $\pm$ 0.047049 & 0.949828 $\pm$ 0.035122\\[3pt]
\hline
12 th & 0.646328 $\pm$ 0.039668 & 0.920086 $\pm$ 0.019238 & 0.198986 $\pm$ 0.023012 & 0.960192 $\pm$ 0.034854\\[3pt]
\hline
13 th & 0.697512 $\pm$ 0.025461 & 0.890253 $\pm$ 0.022778 & 0.201406 $\pm$ 0.030348 & 0.960534 $\pm$ 0.033730\\[3pt]
\hline
14 th & 0.703224 $\pm$ 0.019900 & 0.902898 $\pm$ 0.019830 & 0.205575 $\pm$ 0.038635 & 0.932755 $\pm$ 0.033616\\[3pt]
\hline
15 th & 0.692050 $\pm$ 0.027536 & 0.891654 $\pm$ 0.018239 & 0.200860 $\pm$ 0.024085 & 0.929006 $\pm$ 0.031675\\[3pt]
\hline
16 th & 0.654609 $\pm$ 0.028141 & 0.928181 $\pm$ 0.013477 & 0.202355 $\pm$ 0.017180 & 0.963143 $\pm$ 0.033232\\[3pt]
\hline
17 th & 0.683020 $\pm$ 0.025644 & 0.890392 $\pm$ 0.023586 & 0.198464 $\pm$ 0.033714 & 0.941025 $\pm$ 0.034951\\[3pt]
\hline
18 th & 0.699170 $\pm$ 0.025487 & 0.911382 $\pm$ 0.027290 & 0.188802 $\pm$ 0.036537 & 0.950874 $\pm$ 0.028669\\[3pt]
\hline
19 th & 0.655108 $\pm$ 0.031857 & 0.884214 $\pm$ 0.027816 & 0.209234 $\pm$ 0.032466 & 0.962966 $\pm$ 0.034187\\[3pt]
\hline
20 th & 0.637446 $\pm$ 0.032562 & 0.873755 $\pm$ 0.036792 & 0.196807 $\pm$ 0.026001 & 0.952410 $\pm$ 0.038702\\[3pt]
\hline
\end{tabular}

@@

%%%%%%%%%%%%%%%%
%\subsection{Sensitivity Analysis}

% -- NOT POSSIBLE --

>>>>

% first the title is needed
\title{A Discrepancy-based Framework to Compare Robustness between Multi-Attribute Evaluations}

@@

\begin{abstract}
Multi-objective evaluation is a necessary aspect when managing complex systems, as the intrinsic complexity of a system is generally closely linked to the potential number of optimization objectives.
%In the case of spatial socio-technical systems (territorial and urban systems), the recent example of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where such an evaluation is central to the development of sustainable innovative concepts.
However, an evaluation makes no sense without its robustness being given (in the sense of its reliability). Statistical robustness computation methods are highly dependent of underlying statistical models. We propose a formulation of a model-independent framework in the case of integrated aggregated indicators (Multi-attribute evaluation), that allows to define a relative measure of robustness taking into account data structure and indicator values. We implement and apply it to a synthetic case of urban systems based on Paris districts geography, and to real data for evaluation of income segregation for Greater Paris metropolitan area. First numerical results show the potentialities of this new method. Furthermore, its relative independence to system type and system model may position it as an alternative to classical statistical robustness methods.
\keywords{Multi-attribute Evaluation, Model-Independent Robustness, Urban System, Discrepancy}
\end{abstract}

@@

Multi-objective problems are organically linked to the complexity of underlying systems. Indeed, either in the field of \emph{Complex Industrial Systems}, in the sense of engineered systems, where construction of Systems of Systems (SoS) by coupling and integration often leads to contradictory objectives~\cite{marler2004survey}, or in the field of \emph{Natural Complex Systems}, in the sense of non ingeneered  physical, biological or social systems that exhibit emergence and self-organization properties, where objectives can e.g. be the result of heterogeneous interacting agents (see~\cite{newman2011complex} for a large survey of systems concerned by this approach), multi-objective optimization can be explicitly introduced to study or design the system but is often already implicitly ruling the internal mechanisms of the system. The case of socio-technical Complex Systems is particularly interesting as, following~\cite{haken2003face}, they can be seen as hybrid systems embedding social agents into ``technical artifacts'' (sometimes to an unexpected degree creating what \noun{Picon} describes as \emph{cyborgs}~\cite{picon2013smart}), and thus cumulate propensity to be at the origin of multi-objective issues\footnote{We design by \emph{Multi-Objective Evaluation} all practices including the computation of multiple indicators of a system (it can be multi-objective optimization for system design or multi-objective evaluation of an existing system ; our examples being in the second case but the overall method staying rather generic).}. The new notion of \emph{eco-districts}~\cite{souami2012ecoquartiers} is a typical example where sustainability implies contradictory objectives. The example of transportation systems, which conception shifted during the second half of the 20th century from cost-benefit analysis to multi-criteria decision-making, is also typical of such systems~\cite{bavoux2005geographie}. Geographical system are now well studied from such a point of view in particular thanks to the integration of multi-objective frameworks within Geographical Information Systems~\cite{carver1991integrating}. As for the micro-case of eco-districts, meso and macro urban planning and design may be made sustainable through indicators evaluation~\cite{jegou2012evaluation}.

A crucial aspect of an evaluation is a certain notion of its reliability, that we call here \emph{robustness}.% Various definitions of robustness are possible in different frames, and it will have a precise definition in our framework.
Statistics naturally include this notion since the construction and estimation of statistical models give diverse indicators of the consistence of results~\cite{launer2014robustness}. The first example that comes to mind is the application of the law of large numbers to obtain the \emph{p-value} of a model fit, that can be interpreted as a confidence measure of estimates. Besides, confidence intervals and \emph{beta-power} are other important indicators of statistical robustness. Bayesian inference provide also measures of robustness when distribution of parameters are sequentially estimated. Concerning multi-objective optimization, in particular through heuristic algorithms (for example genetic algorithms, or operational research solvers), the notion of robustness of a solution concerns more the stability of the solution on the phase space of the corresponding dynamical system. Recent progresses have been done towards unified formulation of robustness for a multi-objective optimization problem, such as~\cite{deb2006introducing} where robust Pareto-front as defined as solutions that are insensitive to small perturbations. In~\cite{1688537}, the notion of degree of robustness is introduced, formalized as a sort of continuity of other solutions in successive neighborhood of a solution.

However, there still lack generic methods to estimate robustness of an evaluation that would be model-independent, i.e. that would be extracted from data structure and indicators but that would not depend on the method used. Some advantages could be for example an \emph{a priori} estimation of potential robustness of an evaluation and thus to decide if the evaluation is worth doing. We propose here a framework answering this issue in the particular case of Multi-attribute evaluations, i.e. when the problem is made unidimensional by objectives aggregation. It is data-driven and not model-driven in the sense that robustness estimation does not depend on how indicators are computed, as soon as they respect some assumptions that will be detailed in the following.

@@

%Our approach needs to restrict the scope of application.
We assume that objectives can be expressed as spatial integrals, so it should apply to any territorial system and our application cases are urban systems. It is not that restrictive in terms of possible indicators if one uses suitable variables and integrated kernels : in a way analog to the method of geographically weighted regression~\cite{brunsdon1998geographically}, any spatial variable can be integrated against regular kernels of variable size and the result will be a spatial aggregation which sense depends on kernel size. The example we use in the following such as conditional means or sums suit well the assumption. Even an already spatially aggregated indicator can be interpreted as a spatial indicator by using a Dirac distribution on the centroid of the corresponding area.

@@

A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives, i.e. that we are tackling a multi-attribute optimization problem. If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution. It is analog to aggregated utility techniques in economics and is used in many fields.
%many domains apply such a method that can be called \emph{Multi-Criteria Decision Analysis}~\cite{wang2009review}.
The subtlety lies in the choice of weights, i.e. the shape of the projection function, and various approaches have been developed to find weights depending on the nature of the problem. Recent work~\cite{dobbie2013robustness} proposed to compare robustness of different aggregation techniques through sensitivity analysis, performed by Monte-Carlo simulations on synthetic data. Distribution of biases where obtained for various techniques and some showed to perform significantly better than others. Robustness assessment still depended on models used in that work.

@@

The rest of the paper is organized as follows : section 2 describes intuitively and mathematically the proposed framework ; section 3 then details implementation, data collection for case studies and numerical results for an artificial intra-urban case and a metropolitan real case ; section 4 finally discuss limitations and potentialities of the method.

@@

We describe now the abstract framework allowing theoretically to compare robustnesses of evaluations of two different urban systems. Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context. Intuitively, it relies on empirical base resulting from the following axioms :
\begin{itemize}
\item Urban systems can be seen from the information available, i.e. raw data describing the system. As a data-driven approach, this raw data is the basis of our framework and robustness will be determined by its structure.
\item From data are computed indicators (objective functions). We assume that a choice of indicators is a will to translate particular aspects of the system, that is to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
\item Given many systems and associated indicators, a common space can be built to compare themselves. In that space, data will represent more or less well real systems, depending e.g. on initial scale, precision of data, missing data. We precisely capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory which express how a dataset is well distributed in the space it is embedded in~\cite{dick2010digital}. Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
\end{itemize}

Synthesizing these requirements, we propose a notion of \emph{Robustness} of an evaluation that captures both, by combining data reliability with relative importance,
\begin{enumerate}
\item \emph{Missing Data} : an evaluation based on more refined datasets will naturally be more robust.
\item \emph{Indicator importance} : indicators with more relative influence will weight more on the total robustness.
\end{enumerate}

%%%%%%%%%%%%%%%%
\subsection{Formal Description}

%%%%%%%%%%%%%%%%
\paragraph{Indicators}

@@

%%%%%%%%%%%%%%%%
\paragraph{Aggregation}

%As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation. Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method. 
Weighting objectives in multi-attribute decision-making is indeed the crucial point of the processes, and numerous methods are available (see~\cite{wang2009review} for a review for the particular case of sustainable energy management). Let define weights for the linear aggregation. We assume the indicators normalized, i.e. $q_c \in [0,1]$, for a more simple construction of relative weights. For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is simply constituted by the relative importance of the indicator $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e. the effectively calculated value). Note that this step can be extended to any sets of weight attributions, by taking for example $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ if $\mathbf{w}'$ are the weights attributed by the decision-maker. We focus here on the relative influence of attributes and thus choose this simple form for weights.

%  subjective part is shit.
% subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight. In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.

%%%%%%%%%%%%%%%%
\paragraph{Robustness Estimation}

@@

The intuitive sense of this definition is that one compares robustness of evaluations by comparing the highest error done in each based on data structure and relative importance.

@@

%We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework. It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.

%%%%%%%%%%%%%%%%
\paragraph{Implementation}

@@

%%%%%%%%%%%%%%%%
\subsection{Implementation on Synthetic Data}

We propose in a first stage to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.

%%%%%%%%%%%%%%%%
\paragraph{Data Collection}

We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris. We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features. We use the street network and position of buildings within the city of Paris\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}. Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source. We use centroids of buildings polygons, and segments of street network. Dataset overall consists of around $200k$ building features and $100k$ road segments.
%An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.

%%%%%%%%%%%%%%%%
%\begin{figure}
%\centering
%\includegraphics[width=\textwidth]{figures/ex_data}
%\caption{Example of a visualization of used data for center area of Paris. We use centroids of buildings which polygons are filled in red, and segments of street network drawn in green. Dataset overall consists of around 200000 building features and 100000 road segments.}
%\label{fig:data_ex}
%\end{figure}
%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
\paragraph{Virtual Cases}

@@

\begin{table}[h!]
\hspace{-1cm}
\begin{tabular}[6pt]{c|c|c|c|c}
Arrdt & $<\bar{d}_w> \pm \sigma (\bar{d}_w)$ & $<\bar{\varphi}> \pm \sigma (\bar{\varphi})$ & $<\bar{p}> \pm \sigma (\bar{p})$ & $R_{i,1}$ \\[3pt]
\hline
1 th & 0.731655 $\pm$ 0.041099 & 0.917462 $\pm$ 0.026637 & 0.191615 $\pm$ 0.052142 & 1.000000 $\pm$ 0.000000\\[3pt]
\hline
2 th & 0.723225 $\pm$ 0.032539 & 0.844350 $\pm$ 0.036085 & 0.209467 $\pm$ 0.058675 & 1.002098 $\pm$ 0.039972\\[3pt]
\hline
3 th & 0.713716 $\pm$ 0.044789 & 0.797313 $\pm$ 0.057480 & 0.185541 $\pm$ 0.065089 & 0.999341 $\pm$ 0.048825\\[3pt]
\hline
4 th & 0.712394 $\pm$ 0.042897 & 0.861635 $\pm$ 0.030859 & 0.201236 $\pm$ 0.044395 & 0.973045 $\pm$ 0.036993\\[3pt]
\hline
5 th & 0.715557 $\pm$ 0.026328 & 0.894675 $\pm$ 0.020730 & 0.209965 $\pm$ 0.050093 & 0.963466 $\pm$ 0.040722\\[3pt]
\hline
6 th & 0.733249 $\pm$ 0.026890 & 0.875613 $\pm$ 0.029169 & 0.206690 $\pm$ 0.054850 & 0.990676 $\pm$ 0.031666\\[3pt]
\hline
7 th & 0.719775 $\pm$ 0.029072 & 0.891861 $\pm$ 0.026695 & 0.209265 $\pm$ 0.041337 & 0.966103 $\pm$ 0.037132\\[3pt]
\hline
8 th & 0.713602 $\pm$ 0.034423 & 0.931776 $\pm$ 0.015356 & 0.208923 $\pm$ 0.036814 & 0.973975 $\pm$ 0.033809\\[3pt]
\hline
9 th & 0.712441 $\pm$ 0.027587 & 0.910817 $\pm$ 0.015915 & 0.202283 $\pm$ 0.049044 & 0.971889 $\pm$ 0.035381\\[3pt]
\hline
10 th & 0.713072 $\pm$ 0.028918 & 0.881710 $\pm$ 0.021668 & 0.210118 $\pm$ 0.040435 & 0.991036 $\pm$ 0.038942\\[3pt]
\hline
11 th & 0.682905 $\pm$ 0.034225 & 0.875217 $\pm$ 0.019678 & 0.203195 $\pm$ 0.047049 & 0.949828 $\pm$ 0.035122\\[3pt]
\hline
12 th & 0.646328 $\pm$ 0.039668 & 0.920086 $\pm$ 0.019238 & 0.198986 $\pm$ 0.023012 & 0.960192 $\pm$ 0.034854\\[3pt]
\hline
13 th & 0.697512 $\pm$ 0.025461 & 0.890253 $\pm$ 0.022778 & 0.201406 $\pm$ 0.030348 & 0.960534 $\pm$ 0.033730\\[3pt]
\hline
14 th & 0.703224 $\pm$ 0.019900 & 0.902898 $\pm$ 0.019830 & 0.205575 $\pm$ 0.038635 & 0.932755 $\pm$ 0.033616\\[3pt]
\hline
15 th & 0.692050 $\pm$ 0.027536 & 0.891654 $\pm$ 0.018239 & 0.200860 $\pm$ 0.024085 & 0.929006 $\pm$ 0.031675\\[3pt]
\hline
16 th & 0.654609 $\pm$ 0.028141 & 0.928181 $\pm$ 0.013477 & 0.202355 $\pm$ 0.017180 & 0.963143 $\pm$ 0.033232\\[3pt]
\hline
17 th & 0.683020 $\pm$ 0.025644 & 0.890392 $\pm$ 0.023586 & 0.198464 $\pm$ 0.033714 & 0.941025 $\pm$ 0.034951\\[3pt]
\hline
18 th & 0.699170 $\pm$ 0.025487 & 0.911382 $\pm$ 0.027290 & 0.188802 $\pm$ 0.036537 & 0.950874 $\pm$ 0.028669\\[3pt]
\hline
19 th & 0.655108 $\pm$ 0.031857 & 0.884214 $\pm$ 0.027816 & 0.209234 $\pm$ 0.032466 & 0.962966 $\pm$ 0.034187\\[3pt]
\hline
20 th & 0.637446 $\pm$ 0.032562 & 0.873755 $\pm$ 0.036792 & 0.196807 $\pm$ 0.026001 & 0.952410 $\pm$ 0.038702\\[3pt]
\hline
\end{tabular}

@@

%%%%%%%%%%%%%%%%
\subsection{Application to a Real Case : Metropolitan Segregation}

The first example was aimed to show potentialities of the method but was purely synthetic, hence yielding no concrete conclusion nor implications for policy. We propose now to apply it to real data for the example of metropolitan segregation.

\paragraph{Data}

We work on income data available for France at an intra-urban level (basic statistical units IRIS) for the year 2011 under the form of summary statistics (deciles if the area is populated enough to ensure anonymity), provided by INSEE\footnote{\texttt{http://www.insee.fr}}. Data are associated with geographical extent of statistical units, allowing computation of spatial analysis indicators.

\paragraph{Indicators}

We use here three indicators of segregation integrated on a geographical area. Let assume the area divided into covering units $\mathcal{S}_i$ for $1\leq i \leq N$ with centroids $(x_i,y_i)$. Each unit has characteristics of population $P_i$ and median income $X_i$. We define spatial weights used to quantify strength of geographical interactions between units $i,j$, with $d_{ij}$ euclidian distance between centroids : $w_{ij} = \frac{P_i P_j}{\left(\sum_k P_k\right)^2}\cdot \frac{1}{d_{ij}}$ if $i\neq i$ and $w_{ii} = 0$. The normalized indicators are the following

\begin{itemize}
\item Spatial autocorrelation Moran index, defined as weighted normalized covariance of median income by
\[
\rho = \frac{N}{\sum_{ij}w_{ij}}\cdot \frac{\sum_{ij}w_{ij}\left(X_i - \bar{X}\right)\left(X_j - \bar{X}\right)}{\sum_i \left(X_i - \bar{X}\right)^2}
\]
\item Dissimilarity index (close to Moran but integrating local dissimilarities rather than correlations), with $\tilde{X}_i = \frac{X_i - \min(X_k)}{\max(X_k) - \min(X_k)}$
\[
d =  \frac{1}{\sum_{ij}w_{ij}} \sum_{ij} w_{ij} \left|\tilde{X}_i - \tilde{X}_j\right|
\]
\item Complementary of the entropy of income distribution that is a way to capture global inequalities
\[
\varepsilon = 1 + \frac{1}{\log(N)} \sum_i \frac{X_i}{\sum_k X_k} \cdot \log\left(\frac{X_i}{\sum_k X_k}\right)
\]
\end{itemize}

Numerous measures of segregation with various meanings and at different scales are available, as for example at the level of the unit by comparison of empirical wage distribution with a theoretical null model~\cite{louf2015patterns}. The choice here is arbitrary in order to illustrate our method with a reasonable number of dimensions.

%%%%%%%%%%%%%%%%
\begin{figure}[h!]
\centering
%\vspace{-1.5cm}
\hspace{-3cm}\includegraphics[width=1.2\textwidth]{figures/grandParis_income_moran}
\vspace{-2.5cm}
\caption{\textbf{Maps of Metropolitan Segregation.} Maps show yearly median income on basic statistical units (IRIS) for the three departments constituting mainly the Great Paris metropolitan area, and the corresponding local Moran spatial autocorrelation index, defined for unit $i$ as $\rho_i = N/\sum{j}w_{ij} \cdot \frac{\sum_j w_ij (X_j - \bar{X})(X_i - \bar{X})}{\sum_i (X_i - \bar{X})^2}$. The most segregated areas coincide with the richest and the poorest, suggesting an increase of segregation in extreme situations.}
\end{figure}
%%%%%%%%%%%%%%%%

\paragraph{Results}

We apply our method with these indicators on the Greater Paris area, constituted of four \emph{d{\'e}partements} that are intermediate administrative units. The recent creation of a new metropolitan governance system~\cite{gilli2009paris} underlines interrogations on its consistence, and in particular on its relation to intermediate spatial inequalities. We show in Fig. 1 maps of spatial distribution of median income and corresponding local index of autocorrelation. We observe the well-known West-East opposition and district disparities inside Paris as they were formulated in various studies, such as~\cite{guerois2009dynamique} through the analysis of real estate transactions dynamics. We then apply our framework to answer a concrete question that has implications for urban policy : \textit{how are the evaluation of segregation within different territories sensitive to missing data ?} To do so, we proceed to Monte Carlo simulations (50 repetitions) during which a fixed proportion of data is randomly removed, and the corresponding robustness index is evaluated. Simulations are done on each \emph{department} separately, each time relatively to the robustness of the evaluation of full Greater Paris. Results are shown in Fig. 2. One area (92) present a slightly better robustness than the reference, what could be explained by its homogeneity and thus highly fiable segregation values. Implications for policy that can be drawn are for example direct comparisons between areas : a loss of 8\% of information on Paris area yields an evaluation as robust as with all data in 93 area. The latter being a deprived area, the inequality is increased by this relative lower quality of statistical information. The study of standard deviations suggest further investigations as different response regimes to data removal seem to exist.
